---
title: "Simulating and Calibrating an SIR Model"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate and Calibrate SIR Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette introduces the `simulate_calibrate_sir` function, which simulates and calibrates an SIR (Susceptible-Infected-Recovered) model using TensorFlow and Keras in R. The function is designed to generate simulated epidemic data, train a Convolutional Neural Network (CNN) model, and evaluate the model's performance using the simulated data.

The vignette guides you through the different stages of the function, including data generation, simulation, model building, and evaluation.

## Prerequisites

Ensure that the following libraries are installed and loaded:
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE, # Globally disables evaluation of all chunks
  echo = TRUE   # Shows code in the vignette but does not run it
)
```



```{r}
library(epiworldR)
library(keras3)
library(data.table)
library(tensorflow)
library(reticulate)
reticulate::import("numpy")
library(epiworldRcalibrate)
```

# Function Overview

The `simulate_calibrate_sir` function performs the following key tasks:

1.  **Generate Theta and Seeds:** Randomly generate model parameters and simulation seeds.
2.  **Run Simulations:** Simulate epidemic trajectories based on the generated parameters.
3.  **Prepare Data for Training:** Organize the simulated data to prepare it for training a CNN model.
4.  **Split Data:** Split the data into training and testing sets.
5.  **Build and Train the Model:** Build a CNN model and train it on the training data.
6.  **Evaluate the Model:** Evaluate the trained model's performance on the test data.

The function's arguments are:

-   `N`: Number of simulation runs.
-   `n`: Number of population.
-   `ndays`: Number of days to run the simulation.
-   `ncores`: Number of cores used for parallel processing.

# Step-by-Step Walkthrough

## 1. Generate Theta and Seeds

The first step in the function is to generate the parameter sets (`theta`) and random seeds for the simulation:

```{r}
N=10
n=5000
theta <- generate_theta(N, n)
head(theta,5)


```

```{r}
seeds <- sample.int(.Machine$integer.max, N, TRUE)
head(seeds,5)
```

-   `generate_theta()`: Generates random parameters for the SIR model.
-   `seeds`: Random seeds are generated to ensure reproducibility of the simulations.

## 2. Run Simulations

Next, the function runs the simulations for the SIR model using the generated parameters and seeds and using epiworldR package:

```{r}
N=10
n=5000
ndays=50
ncores=20
matrices <- run_simulations(N, n, ndays, ncores, theta, seeds)
matrices[[1]][1,,] 
```

-   The `run_simulations()` function generates and saves the simulated epidemic trajectories to the specified path.

## 3. Filter Non-Null Data

After running the simulations, the function filters out incomplete or null elements to ensure data quality:

```{r}
filtered_data <- filter_non_null(matrices, theta)
matrices <- filtered_data$matrices
head(matrices[[1]][1,,])
```

```{r}
theta <- filtered_data$theta
N <- filtered_data$N
N
```

N is the number of non null datasets that we could simulate. - `filter_non_null()`: Filters out any null simulations and updates `theta` and `N` accordingly.

## 4. Prepare Data for TensorFlow

To use the simulated data for training, the matrices are prepared and converted into arrays that can be processed by TensorFlow:

```{r}
arrays_1d <- prepare_data_for_tensorflow(matrices, N)
head(arrays_1d[,,1])
# Save theta and simulations data
  theta2 <-as.data.table(copy(theta))
  theta2$crate <- plogis(theta2$crate / 10)
```

-   `prepare_data_for_tensorflow()`: Reshapes the simulated data into the format required for TensorFlow training.

The transformation is essential because contact rate (crate) originally has values greater than 0, and the CNN expects normalized values between 0 and 1 for effective training. By scaling (crate / 10) and then applying the logistic transformation (plogis()), we ensure that the contact rate is expressed as a probability, which is easier for the CNN to handle and learn from. \## 5. Split Data into Training and Testing Sets

The dataset is split into training and testing sets to evaluate the model's performance: (our default is to use 70% of the data as train and 30% as test)

```{r}
data_split <- split_data(arrays_1d, theta2, N)
  train <- data_split$train
  test <- data_split$test
```

-   `split_data()`: Splits the data into `train` and `test` sets.

## 6. Build and Train the CNN Model

The Convolutional Neural Network (CNN) model is built and trained on the simulation data:

```{r}
model <- build_cnn_model(dim(arrays_1d)[-1], ncol(theta))
train_model(model, train, epochs=2, verbose=1)
```

-   `build_cnn_model()`: Builds a CNN model with input dimensions based on the simulation data.
-   `train_model()`: Trains the model using the training dataset.
    -   `epochs`: The number of times the learning algorithm will work through the entire training dataset.
    -   `verbose`: Controls the level of logging output during training, where `0` is silent, `1` shows progress bars, and `2` displays one line per epoch.

## 7. Evaluate the Model

Finally, the function evaluates the performance of the trained model on the test data:

```{r}
eval_results <- evaluate_model(model, test, theta)
pred <- eval_results$pred
MAEs <- eval_results$MAEs
```

-   `evaluate_model()`: Evaluates the model's performance and calculates Mean Absolute Errors (MAEs).
-   `pred`: Predictions made by the model.
-   `MAEs`: Mean Absolute Errors between the predicted and actual parameter values.

```{r}
print(head(pred,10))
```

```{r}
print(MAEs)
```

These was all the steps behind the `simulate_calibrate_sir` function but you can just call this function and see results:(it may take 2 minutes)

```{r}
N=2e4
n=5000
ndays=50
epochs=100
ncores=20
verbose=2
simulate_calibrate_sir(N,n,ndays,ncores,epochs,verbose)
```

# Conclusion

The `simulate_calibrate_sir` function provides a comprehensive workflow to generate, simulate, and calibrate an SIR model using deep learning. It leverages TensorFlow and Keras to create a powerful calibration model, and it offers valuable insights into how well a CNN can learn the dynamics of a simulated epidemic.

The workflow covers all aspects of data simulation, processing, and model training. By following this vignette, you should be able to understand the full pipeline for simulating and calibrating an SIR model.

Feel free to modify and extend the function based on your specific requirements.

# Section 2: See the Calibration

After running the simulation and calibration of the SIR model, it is essential to examine the calibration results to understand how well the model performed.

The following steps can be taken to see the calibration:

**Predict Parameters Using CNN Model:** After training the Convolutional Neural Network (CNN) model, you can use the `calibrate_sir(data)` function to predict the parameters of the SIR model based on input data. The `data` parameter should be a numeric matrix or array containing counts of infected individuals over a period of 30 or 60 days.

```{r}
N=1
n=5000
set.seed(123)
theta <- generate_theta(N, n)
ncores=20
ndays=60
seeds=123
theta[1,2]=7
# Run simulations
i=1
m <- epiworldR:: ModelSIRCONN(
  "mycon",
  prevalence        = 0.12,
  contact_rate      = 6.09,
  transmission_rate = 0.1,
  recovery_rate     = 1/7,
  n                 = n
)
ndays=30
plogis(6.09/10)
verbose_off(m)
run(m, ndays = ndays)
incidence = epiworldR::plot_incidence(m, plot = FALSE)
data=incidence$Infected
length(data)
# Save theta and simulations data
theta2 <-as.data.table(copy(theta))
theta2$crate <- plogis(theta2$crate / 10)
result=calibrate_sir(data)
```

The output of `calibrate_sir` will include predicted values such as `preval` (prevalence), `crate` (case rate), `ptran` (transmission probability), and `prec` (precision).

**Evaluate Calibration Accuracy:** To evaluate how well the calibration has worked, you can compare the predicted parameters with the true parameters that were used to generate the simulation. You can use metrics such as Mean Absolute Error (MAE) or Root Mean Square Error (RMSE) to measure the accuracy of the predictions.

```{r}
# Assuming you have a set of true parameters for comparison
predicted <- result$pred  # Replace with actual column name for predictions
actual <- (data.frame(preval=0.12,crate=0.64,ptran=0.1,prec=0.14))  # Assuming 'data' contains the actual infected numbers
MAE=abs((predicted-actual))
# Combine actual and predicted values into a data table
comparison <- data.frame(rbind(
  Actual = actual,
  Predicted = predicted
))
rownames(comparison)=c("Actual", "Predicted")
print(comparison)
```
Now, I want to compare the plots of SIR model with predicted values and actual values:
```{r}
# Apply reverse logistic transformation to theta2$crate / 10
predicted$new <- 6.63
plogis(7/10)
m2 <- epiworldR:: ModelSIRCONN(
  "mycon",
  prevalence        = 0.02,
  contact_rate      = 7,
  transmission_rate = 0.27,
  recovery_rate     = 0.49,
  n                 = n
)

verbose_off(m2)
run(m2, ndays = 30)
par(mfrow = c(1, 2))
plot(m2, main = "SIR Predicted Parameters")
```


```{r}
plot(m, main = "SIR Actual Parameters")

```


```{r}
incidence1 = epiworldR::plot_incidence(m, plot = TRUE)
```


```{r}
incidence2 = epiworldR::plot_incidence(m2, plot = TRUE)
plot(incidence1$Infected)
plot(incidence2$Infected)

# Extract data from the incidence objects
# Assuming incidence1 and incidence2 are already created using plot_incidence()
incidence1_data <- data.frame(Days = 1:nrow(incidence1), Infected = incidence1$Infected, Type = "Actual")
incidence2_data <- data.frame(Days = 1:nrow(incidence2), Infected = incidence2$Infected, Type = "Predicted")

# Combine the data for plotting
combined_data <- rbind(incidence1_data, incidence2_data)

# Load necessary library
library(ggplot2)
```


```{r}
# Create the plot
ggplot(combined_data, aes(x = Days, y = Infected, color = Type, linetype = Type)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("Predicted" = "darkred", "Actual" = "blue")) +
  labs(
    title = "Infection Incidence: Actual vs Predicted",
    x = "Days",
    y = "Number of Infected Individuals",
    color = "Legend",
    linetype = "Legend"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )


```

```{r}
# Plot the predicted vs actual values
library(ggplot2)
library(ggrepel)
ggplot(comparison, aes(x = Actual, y = Predicted, label = Parameter)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(color = Parameter), size = 4) +
  geom_text_repel(size = 4, max.overlaps = 20) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Calibration of SIR Model Parameters",
    subtitle = "Comparing Predicted vs. Actual Values",
    x = "Actual Values",
    y = "Predicted Values",
    color = "Parameter"
  ) +
  theme_light(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  )
comparison <- data.frame(
  Parameter = c("Prevalence", "Contact Rate", "Transmission", "Recovery Rate"),
  Predicted = unlist(predicted),
  Actual = unlist(actual)
)



```

# References

-   [Keras Documentation](https://keras.rstudio.com/)
-   [TensorFlow for R](https://tensorflow.rstudio.com/)
-   [Epidemiology Models in R](https://CRAN.R-project.org/)
-   [epiworld on GitHub](https://github.com/UofUEpiBio/epiworld)

# Session Info

```{r}
sessionInfo()

```

```{r}
set.seed(123)
theta=generate_theta(1,5000)

m <- epiworldR:: ModelSIRCONN(
  "mycon",
  prevalence        = 0.119,
  contact_rate      = 6.09,
  transmission_rate = 0.1,
  recovery_rate     = 1/7,
  n                 = 5000
)

run(m, ndays = 61)
incidence = epiworldR::plot_incidence(m, plot = FALSE)
data=incidence$Infected

# Save theta and simulations data
theta2 <-as.data.table(copy(theta))
theta2$crate <- plogis(theta2$crate/10)



result=calibrate_sir(data)
data <- data.frame(
  Category = c("Actual", "Predicted", "MAE"),
  preval = c(0.1190000, 0.2230902, 0.1040902),
  crate = c(0.64784451, 0.61379176, 0.03405274),
  ptran = c(0.0797096, 0.2127937, 0.1330841),
  prec = c(0.54575686, 0.53561789, 0.01013897)
)

# Calculate adjusted_crate using the formula plogis(x/10) = crate
data$adjusted_crate <- log(data$crate / (1 - data$crate)) * 10
data$adjusted_crate[3]=1.5
# Print the table
print(data)
6.1-4.6
```
```{r}

```

