---
title: "Getting started with epiworldRcalibrate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with epiworldRcalibrate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Overview

`epiworldRcalibrate` separates the epidemic model from the calibration method.

The package uses the same three-step workflow for every model and method:

```text
1. simulation/data -> 2. training/calibrator -> 3. calibrate/test/compare
```

The three main R files follow the same organization:

```text
R/01_simulation.R
R/02_training.R
R/03_test_compare.R
```

# The calibrator interface

A calibration method is represented by a simple R object with three functions:

```{r}
calibrator_example <- list(

  pre_process = function(x, ...) {
    x
  },

  run = function(x, ...) {
    x
  },

  post_process = function(x, ...) {
    x
  },

  type = "SIR"

) |>
  structure(class = "calibrator")

calibrator_example
```

All calibration methods are called through the same function. For this
identity example, use a small numeric curve:

```{r}
example_observed_data <- c(1, 2, 3, 4, 5)

calibrate(
  daily_cases = example_observed_data,
  calibrator = calibrator_example
)
```

Real calibrators can receive named known parameters such as `n = 8000` and
`recov = 0.1`.

Internally, `calibrate()` simply runs:

```text
pre_process -> run -> post_process
```

A new deep-learning model, Bayesian method, optimizer, or other method only
needs to follow this `calibrator` interface.

# Step 1: create or import training data

## Built-in SIR

The package can generate SIR training data with `epiworldR`:

```{r}
sir_data <- generate_sir_training_data(
  n_sims = 50,
  ndays = 60,
  seed = 122,
  save = TRUE
)

head(sir_data$theta)
dim(sir_data$observations)
```

For a real training run, increase `n_sims` substantially. The small value above
is only for demonstrating the workflow.

## Built-in SEIR

The package can also generate SEIR training data:

```{r}
seir_data <- generate_seir_training_data(
  n_sims = 20,
  ndays = 365,
  seed = 122,
  save = TRUE
)

head(seir_data$theta)
dim(seir_data$observations)
```

Again, the small number of simulations is only for a quick example.

## A model not available in epiworldR

For a model such as SIRH or SEIRD, the user can supply their own training bank.

The parameter table must contain one row per simulation:

```{r}
my_parameter_table <- data.frame(
  n = c(5000, 6000, 7000),
  gamma = c(0.10, 0.12, 0.09),
  hosp_rate = c(0.05, 0.08, 0.06),
  beta = c(0.25, 0.30, 0.20)
)
```

The observation matrix must contain one observed time series per row:

```{r}
my_observation_matrix <- rbind(
  c(1, 3, 5, 7, 6),
  c(2, 4, 8, 9, 5),
  c(1, 2, 4, 6, 3)
)
```

The observations do not have to be case incidence. They can represent
hospitalizations, deaths, prevalence, or another numeric time series.

```{r}
custom_data <- new_training_data(
  parameters = my_parameter_table,
  observations = my_observation_matrix,
  type = "SIRH"
)

custom_data
```

This example only demonstrates the required data shape. In a real application,
the user would normally provide many simulated parameter/observation pairs.

If the user has their own simulator rather than an existing training bank,
they can connect it through `new_simulator()` and then call
`simulate_training_data()`.

# Step 2: train or define a calibrator

## Train the generic BiLSTM

For SIR:

```{r}
train_bilstm_calibrator(
  name = "sir",
  known_names = c("n", "recov"),
  target_names = c("ptran", "crate", "R0"),
  type = "SIR",
  package_dir = find_project_root(),
  epochs = 100,
  physics_weight = 0.03
)
```

The fitted model and preprocessing objects are saved under:

```text
inst/models/sir/trained/
```

Create the SIR BiLSTM calibrator:

```{r}
sir_bilstm <- bilstm_calibrator(
  model_dir = "inst/models/sir/trained",
  known_names = c("n", "recov"),
  type = "SIR"
)

sir_bilstm
```

The returned object is still just a standard `calibrator`:

```text
pre_process
run
post_process
type
```

## Use the original pretrained SIR model

The original SIR model can be used without retraining:

```{r}
sir_pretrained <- pretrained_sir_calibrator()

sir_pretrained
```

## Define a completely new calibration method

A user can add another method without changing `calibrate()`:

```{r}
calibrator_custom <- list(

  pre_process = function(x, population_size, ...) {
    list(
      observed = as.numeric(x),
      n = population_size
    )
  },

  run = function(x, ...) {
    # Replace this with the user's calibration algorithm.
    list(
      beta = 0.25,
      n = x$n
    )
  },

  post_process = function(x, ...) {
    unlist(x)
  },

  type = "SIRH"

) |>
  structure(class = "calibrator")
```

# Step 3: calibrate and compare

## Create one SIR curve for testing

First create a simulator:

```{r}
sir_sim <- sir_simulator(ndays = 60)
```

Choose known true parameters:

```{r}
true_parameters <- list(
  n = 8000,
  preval = 0.01,
  recov = 0.10,
  crate = 3,
  R0 = 2.5,
  ptran = 2.5 * 0.10 / 3
)
```

Generate one observed curve:

```{r}
curve <- simulate_from_parameters(
  simulator = sir_sim,
  parameters = true_parameters,
  seed = 3123
)

length(curve)
```

## Calibrate with BiLSTM

```{r}
fit <- calibrate(
  daily_cases = curve,
  calibrator = sir_bilstm,
  n = true_parameters$n,
  recov = true_parameters$recov
)

fit
```

The same `calibrate()` function is used for every calibration method.

# Compare against standard baseline methods

The package includes the four methods used as standard comparisons in this
project:

```text
ABC-MCMC
ABC-SMC
Nelder-Mead
Differential Evolution
```

For example, suppose SIR calibration estimates `crate` and `ptran`:

```{r}
bounds <- list(
  crate = c(1, 5),
  ptran = c(0.001, 1)
)
```

Known parameters are supplied separately:

```{r}
known_names <- c(
  "n",
  "preval",
  "recov"
)
```

Create the four baseline calibrators:

```{r}
baselines <- baseline_calibrators(
  simulator = sir_sim,
  bounds = bounds,
  known_names = known_names,
  type = "SIR"
)
```

A primary calibration method can then be added to the same list:

```{r}
methods <- c(
  list(BiLSTM = sir_bilstm),
  baselines
)
```

The generic comparison function runs all supplied calibrators on the same
observation:

```{r}
comparison <- compare_calibrators(
  observed = curve,
  calibrators = methods,
  known = list(
    n = true_parameters$n,
    preval = true_parameters$preval,
    recov = true_parameters$recov
  ),
  truth = c(
    ptran = true_parameters$ptran,
    crate = true_parameters$crate,
    R0 = true_parameters$R0
  ),
  simulator = sir_sim
)

comparison$summary
comparison$estimates
```

# Summary

The package supports two data-generation situations:

```text
SIR / SEIR
    -> training data can be generated by the package

SIRH / SEIRD / another model
    -> user supplies a training bank
       OR
       user connects their own simulator
```

The calibration interface is the same in every case:

```r
calibrator_x <- list(
  pre_process = function(x, ...) { ... },
  run = function(x, ...) { ... },
  post_process = function(x, ...) { ... },
  type = "MODEL"
) |>
  structure(class = "calibrator")
```

and every method is used through:

```r
calibrate(observed_data, calibrator_x, ...)
```
