---
title: "Run SIR and SEIR BiLSTM Calibration"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = TRUE,
  warning = TRUE
)

# ---------------------------------------------------------------------------
# Find the package root whether this file is run from the project root,
# examples/, or vignettes/.
# ---------------------------------------------------------------------------
find_root <- function() {
  candidates <- c(".", "..", "../..", "../../..")

  for (x in candidates) {
    if (
      file.exists(file.path(x, "DESCRIPTION")) &&
      dir.exists(file.path(x, "R")) &&
      dir.exists(file.path(x, "inst"))
    ) {
      return(normalizePath(x, mustWork = TRUE))
    }
  }

  stop(
    "Could not find the epiworldRcalibrate project root. ",
    "Open the R project and run this file from inside the package."
  )
}

PROJECT_ROOT <- find_root()

# Make relative paths in all later chunks resolve from the package root.
knitr::opts_knit$set(root.dir = PROJECT_ROOT)

# Source the three user-facing R files.
source(file.path(PROJECT_ROOT, "R", "01_simulation.R"))
source(file.path(PROJECT_ROOT, "R", "02_training.R"))
source(file.path(PROJECT_ROOT, "R", "03_test_compare.R"))

# ---------------------------------------------------------------------------
# RUN SETTINGS
#
# FALSE = pipeline test: smaller banks and fewer epochs
# TRUE  = full current research settings
# ---------------------------------------------------------------------------
FULL_RUN <- FALSE
RETRAIN <- TRUE

if (FULL_RUN) {

  SIR_N_SIMS  <- 50L
  SEIR_N_SIMS <- 20L
  EPOCHS      <- 10L

} else {

  # Quick mode is for checking that the full workflow runs successfully.
  # Do NOT use its accuracy values as final scientific results.
  SIR_N_SIMS  <- 10L
  SEIR_N_SIMS <- 10L
  EPOCHS      <- 10L
}

BATCH_SIZE <- 64L
SEED <- 122L

if (!FULL_RUN) {
  message(
    "QUICK MODE: pipeline test only. Accuracy is not expected to be good ",
    "with 100 simulations and 10 epochs."
  )
}

RESULTS_DIR <- file.path(PROJECT_ROOT, "results")
dir.create(RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)
```

# Purpose

This notebook **actually runs** the SIR and SEIR workflows.

For each model it:

```text
1. generates training data with epiworldR
2. saves the training bank
3. trains the generic Python BiLSTM
4. reads validation metrics
5. creates a new test epidemic curve
6. loads the trained model as a calibrator
7. estimates the unknown parameters
8. compares prediction with truth
9. saves the results
```

The reusable Python model used for **both** SIR and SEIR is:

```text
inst/python/bilstm.py
```

The small command-line training wrapper is:

```text
inst/python/train_bilstm.py
```

The R/Python separation is:

```text
R/01_simulation.R
    |
    | epiworldR generates training curves
    v
data/*_theta.csv
data/*_observations.csv
    |
    v
R/02_training.R
    |
    | calls
    v
inst/python/bilstm.py
    |
    v
inst/models/<model>/trained/
    |
    v
R/03_test_compare.R
    |
    v
calibrate()
```

# 0. Check the environment

```{r environment}
cat("Project root:", PROJECT_ROOT, "\n")
cat("Full run:", FULL_RUN, "\n")
cat("Retrain:", RETRAIN, "\n")
cat("SIR simulations:", SIR_N_SIMS, "\n")
cat("SEIR simulations:", SEIR_N_SIMS, "\n")
cat("Epochs:", EPOCHS, "\n\n")

if (!requireNamespace("epiworldR", quietly = TRUE)) {
  stop("The R package 'epiworldR' is required for SIR/SEIR simulation.")
}

if (!requireNamespace("reticulate", quietly = TRUE)) {
  stop("The R package 'reticulate' is required for Python training.")
}

setup_python_deps()

cat("\nPython configuration:\n")
print(reticulate::py_config())
```

# Part I — SIR

## 1. Generate SIR training data with epiworldR

Current SIR parameter ranges are:

```text
n       = 5000 to 10000
recov   ~ Uniform(0.071, 0.25)
preval  ~ Uniform(0.007, 0.02)
crate   ~ Uniform(1, 5)
R0      ~ Uniform(1.1, 5)

ptran = R0 * recov / crate
```

The simulator is `epiworldR::ModelSIRCONN()`.

```{r sir-generate}
set.seed(SEED)

sir_data <- generate_sir_training_data(
  n_sims = SIR_N_SIMS,
  ndays = 60,
  seed = SEED,
  save = TRUE,
  name = "sir",
  package_dir = PROJECT_ROOT
)

cat("SIR parameter rows:", nrow(sir_data$theta), "\n")
cat("SIR curve length:", ncol(sir_data$observations), "\n\n")

head(sir_data$theta)
```

The training files are:

```{r sir-data-files}
sir_theta_file <- file.path(
  PROJECT_ROOT,
  "data",
  "sir_theta.csv"
)

sir_curve_file <- file.path(
  PROJECT_ROOT,
  "data",
  "sir_observations.csv"
)

cat("theta exists:", file.exists(sir_theta_file), "\n")
cat("observations exist:", file.exists(sir_curve_file), "\n")
cat("theta:", sir_theta_file, "\n")
cat("observations:", sir_curve_file, "\n")
```

## 2. Train the SIR BiLSTM

Known inputs:

```text
n
recov
```

Targets:

```text
ptran
crate
R0
```

The mechanistic relationship used by the optional physics loss is:

```text
R0 * recov = ptran * crate
```

```{r sir-train}
sir_model_dir <- file.path(
  PROJECT_ROOT,
  "inst",
  "models",
  "sir",
  "trained"
)

if (RETRAIN || !file.exists(file.path(sir_model_dir, "model.pt"))) {

  train_bilstm_calibrator(
    name = "sir",
    known_names = c(
      "n",
      "recov"
    ),
    target_names = c(
      "ptran",
      "crate",
      "R0"
    ),
    type = "SIR",
    package_dir = PROJECT_ROOT,
    epochs = EPOCHS,
    batch_size = BATCH_SIZE,
    learning_rate = 0.001,
    seed = SEED,
    hidden_size = 160,
    num_layers = 3,
    dropout = 0.5,
    physics_weight = 0.03
  )

} else {

  cat("Using existing SIR model:", sir_model_dir, "\n")
}
```

## 3. SIR validation results

```{r sir-validation}
sir_validation <- read.csv(
  file.path(
    sir_model_dir,
    "validation_metrics.csv"
  )
)

sir_validation$model <- "SIR"

knitr::kable(
  sir_validation,
  digits = 4,
  caption = "SIR BiLSTM validation metrics"
)
```

### SIR training history

```{r sir-training-history, fig.width=7, fig.height=5}
sir_history <- read.csv(
  file.path(
    sir_model_dir,
    "training_history.csv"
  )
)

matplot(
  sir_history$epoch,
  cbind(
    sir_history$train_loss,
    sir_history$validation_loss
  ),
  type = "l",
  lty = 1,
  xlab = "Epoch",
  ylab = "Loss",
  main = "SIR BiLSTM training history"
)

legend(
  "topright",
  legend = c(
    "Training loss",
    "Validation loss"
  ),
  lty = 1,
  bty = "n"
)
```

### SIR true versus predicted validation parameters

```{r sir-validation-scatter, fig.width=6, fig.height=5}
sir_predictions <- read.csv(
  file.path(
    sir_model_dir,
    "validation_predictions.csv"
  )
)

for (parameter in c("ptran", "crate", "R0")) {

  truth <- sir_predictions[[paste0("true_", parameter)]]
  pred  <- sir_predictions[[paste0("pred_", parameter)]]

  plot(
    truth,
    pred,
    pch = 16,
    xlab = paste("True", parameter),
    ylab = paste("Predicted", parameter),
    main = paste("SIR:", parameter)
  )

  abline(
    0,
    1,
    lty = 2
  )
}
```

## 4. Create a new SIR test curve

This test epidemic is **not** one of the training curves.

```{r sir-test-data}
sir_sim <- sir_simulator(
  ndays = 60
)

true_sir <- list(
  n = 8000,
  preval = 0.01,
  recov = 0.10,
  crate = 3,
  R0 = 2.5,
  ptran = 2.5 * 0.10 / 3
)

sir_curve <- simulate_from_parameters(
  simulator = sir_sim,
  parameters = true_sir,
  seed = 3123
)

cat("Test SIR curve length:", length(sir_curve), "\n")
cat("Total SIR incidence:", sum(sir_curve), "\n")
```

```{r sir-test-curve, fig.width=8, fig.height=4}
plot(
  0:60,
  sir_curve,
  type = "l",
  xlab = "Day",
  ylab = "Incidence",
  main = "SIR test epidemic"
)
```

## 5. Calibrate the SIR test curve

```{r sir-load-calibrator}
sir_bilstm <- bilstm_calibrator(
  model_dir = sir_model_dir,
  known_names = c(
    "n",
    "recov"
  ),
  type = "SIR",
  package_dir = PROJECT_ROOT
)

sir_bilstm
```

```{r sir-calibrate}
sir_fit <- calibrate(
  daily_cases = sir_curve,
  calibrator = sir_bilstm,
  n = true_sir$n,
  recov = true_sir$recov
)

sir_fit
```

## 6. SIR truth versus prediction

```{r sir-final-results}
sir_truth <- c(
  ptran = true_sir$ptran,
  crate = true_sir$crate,
  R0 = true_sir$R0
)

sir_result <- data.frame(
  model = "SIR",
  parameter = names(sir_truth),
  truth = as.numeric(sir_truth),
  prediction = as.numeric(
    sir_fit[names(sir_truth)]
  )
)

sir_result$absolute_error <-
  abs(
    sir_result$prediction -
    sir_result$truth
  )

knitr::kable(
  sir_result,
  digits = 5,
  caption = "SIR held-out test result"
)

write.csv(
  sir_result,
  file.path(
    RESULTS_DIR,
    "sir_bilstm_test_result.csv"
  ),
  row.names = FALSE
)
```

# Part II — SEIR

## 7. Generate SEIR training data with epiworldR

Current SEIR parameter ranges are:

```text
n                 = 5000 to 10000
recov             ~ Uniform(0.071, 0.25)
crate             ~ Uniform(5, 15)
incub             ~ Uniform(3, 21)
R0                ~ Uniform(1, 5)
initial_infected  = 100 to 2000

preval = initial_infected / n
ptran  = R0 * recov / crate
```

The simulator is `epiworldR::ModelSEIRCONN()`.

```{r seir-generate}
set.seed(SEED)

seir_data <- generate_seir_training_data(
  n_sims = SEIR_N_SIMS,
  ndays = 365,
  seed = SEED,
  save = TRUE,
  name = "seir",
  package_dir = PROJECT_ROOT
)

cat("SEIR parameter rows:", nrow(seir_data$theta), "\n")
cat("SEIR curve length:", ncol(seir_data$observations), "\n\n")

head(seir_data$theta)
```

The training files are:

```{r seir-data-files}
seir_theta_file <- file.path(
  PROJECT_ROOT,
  "data",
  "seir_theta.csv"
)

seir_curve_file <- file.path(
  PROJECT_ROOT,
  "data",
  "seir_observations.csv"
)

cat("theta exists:", file.exists(seir_theta_file), "\n")
cat("observations exist:", file.exists(seir_curve_file), "\n")
cat("theta:", seir_theta_file, "\n")
cat("observations:", seir_curve_file, "\n")
```

## 8. Train the SEIR BiLSTM

Known inputs:

```text
n
recov
incub
```

Targets:

```text
ptran
crate
R0
```

```{r seir-train}
seir_model_dir <- file.path(
  PROJECT_ROOT,
  "inst",
  "models",
  "seir",
  "trained"
)

if (RETRAIN || !file.exists(file.path(seir_model_dir, "model.pt"))) {

  train_bilstm_calibrator(
    name = "seir",
    known_names = c(
      "n",
      "recov",
      "incub"
    ),
    target_names = c(
      "ptran",
      "crate",
      "R0"
    ),
    type = "SEIR",
    package_dir = PROJECT_ROOT,
    epochs = EPOCHS,
    batch_size = BATCH_SIZE,
    learning_rate = 0.001,
    seed = SEED,
    hidden_size = 160,
    num_layers = 3,
    dropout = 0.5,
    physics_weight = 0.03
  )

} else {

  cat("Using existing SEIR model:", seir_model_dir, "\n")
}
```

## 9. SEIR validation results

```{r seir-validation}
seir_validation <- read.csv(
  file.path(
    seir_model_dir,
    "validation_metrics.csv"
  )
)

seir_validation$model <- "SEIR"

knitr::kable(
  seir_validation,
  digits = 4,
  caption = "SEIR BiLSTM validation metrics"
)
```

### SEIR training history

```{r seir-training-history, fig.width=7, fig.height=5}
seir_history <- read.csv(
  file.path(
    seir_model_dir,
    "training_history.csv"
  )
)

matplot(
  seir_history$epoch,
  cbind(
    seir_history$train_loss,
    seir_history$validation_loss
  ),
  type = "l",
  lty = 1,
  xlab = "Epoch",
  ylab = "Loss",
  main = "SEIR BiLSTM training history"
)

legend(
  "topright",
  legend = c(
    "Training loss",
    "Validation loss"
  ),
  lty = 1,
  bty = "n"
)
```

### SEIR true versus predicted validation parameters

```{r seir-validation-scatter, fig.width=6, fig.height=5}
seir_predictions <- read.csv(
  file.path(
    seir_model_dir,
    "validation_predictions.csv"
  )
)

for (parameter in c("ptran", "crate", "R0")) {

  truth <- seir_predictions[[paste0("true_", parameter)]]
  pred  <- seir_predictions[[paste0("pred_", parameter)]]

  plot(
    truth,
    pred,
    pch = 16,
    xlab = paste("True", parameter),
    ylab = paste("Predicted", parameter),
    main = paste("SEIR:", parameter)
  )

  abline(
    0,
    1,
    lty = 2
  )
}
```

## 10. Create a new SEIR test curve

```{r seir-test-data}
seir_sim <- seir_simulator(
  ndays = 365
)

true_seir <- list(
  n = 8000,
  recov = 0.10,
  crate = 8,
  incub = 7,
  R0 = 2.5,
  initial_infected = 100
)

true_seir$preval <-
  true_seir$initial_infected /
  true_seir$n

true_seir$ptran <-
  true_seir$R0 *
  true_seir$recov /
  true_seir$crate

seir_curve <- simulate_from_parameters(
  simulator = seir_sim,
  parameters = true_seir,
  seed = 4123
)

cat("Test SEIR curve length:", length(seir_curve), "\n")
cat("Total SEIR incidence:", sum(seir_curve), "\n")
```

```{r seir-test-curve, fig.width=8, fig.height=4}
plot(
  seq_len(365),
  seir_curve,
  type = "l",
  xlab = "Day",
  ylab = "Incidence",
  main = "SEIR test epidemic"
)
```

## 11. Calibrate the SEIR test curve

```{r seir-load-calibrator}
seir_bilstm <- bilstm_calibrator(
  model_dir = seir_model_dir,
  known_names = c(
    "n",
    "recov",
    "incub"
  ),
  type = "SEIR",
  package_dir = PROJECT_ROOT
)

seir_bilstm
```

```{r seir-calibrate}
seir_fit <- calibrate(
  daily_cases = seir_curve,
  calibrator = seir_bilstm,
  n = true_seir$n,
  recov = true_seir$recov,
  incub = true_seir$incub
)

seir_fit
```

## 12. SEIR truth versus prediction

```{r seir-final-results}
seir_truth <- c(
  ptran = true_seir$ptran,
  crate = true_seir$crate,
  R0 = true_seir$R0
)

seir_result <- data.frame(
  model = "SEIR",
  parameter = names(seir_truth),
  truth = as.numeric(seir_truth),
  prediction = as.numeric(
    seir_fit[names(seir_truth)]
  )
)

seir_result$absolute_error <-
  abs(
    seir_result$prediction -
    seir_result$truth
  )

knitr::kable(
  seir_result,
  digits = 5,
  caption = "SEIR held-out test result"
)

write.csv(
  seir_result,
  file.path(
    RESULTS_DIR,
    "seir_bilstm_test_result.csv"
  ),
  row.names = FALSE
)
```

# Part III — Combined results

## 13. Validation metrics for both models

```{r combined-validation}
validation_results <- rbind(
  sir_validation,
  seir_validation
)

validation_results <- validation_results[
  ,
  c(
    "model",
    "parameter",
    "MAE",
    "RMSE",
    "R2"
  )
]

knitr::kable(
  validation_results,
  digits = 4,
  caption = "BiLSTM validation results: SIR and SEIR"
)

write.csv(
  validation_results,
  file.path(
    RESULTS_DIR,
    "bilstm_validation_metrics.csv"
  ),
  row.names = FALSE
)
```

## 14. Held-out test predictions for both models

```{r combined-test-results}
test_results <- rbind(
  sir_result,
  seir_result
)

knitr::kable(
  test_results,
  digits = 5,
  caption = "Held-out calibration results: SIR and SEIR"
)

write.csv(
  test_results,
  file.path(
    RESULTS_DIR,
    "bilstm_sir_seir_test_results.csv"
  ),
  row.names = FALSE
)
```

# 15. Files produced by this run

```{r output-files}
cat(
  "
Training data:
  data/sir_theta.csv
  data/sir_observations.csv
  data/seir_theta.csv
  data/seir_observations.csv

SIR trained model:
  inst/models/sir/trained/model.pt
  inst/models/sir/trained/model_config.json
  inst/models/sir/trained/validation_metrics.csv
  inst/models/sir/trained/validation_predictions.csv
  inst/models/sir/trained/training_history.csv

SEIR trained model:
  inst/models/seir/trained/model.pt
  inst/models/seir/trained/model_config.json
  inst/models/seir/trained/validation_metrics.csv
  inst/models/seir/trained/validation_predictions.csv
  inst/models/seir/trained/training_history.csv

Combined results:
  results/bilstm_validation_metrics.csv
  results/sir_bilstm_test_result.csv
  results/seir_bilstm_test_result.csv
  results/bilstm_sir_seir_test_results.csv
"
)
```

# 16. Full research run

The default at the top of this notebook is:

```r
FULL_RUN <- FALSE
```

That is only to verify the pipeline.

For the current full SIR/SEIR experiment, change it to:

```r
FULL_RUN <- TRUE
```

Then the notebook uses:

```text
SIR:
    5,000 simulations
    60 days + day zero = 61 observations
    100 epochs

SEIR:
    20,000 simulations
    365 observations
    100 epochs
```

Those are the settings whose validation metrics should be used for the actual
scientific comparison, not the small quick-run settings.

# 17. Standard baseline comparison

ABC-MCMC, ABC-SMC, Nelder-Mead, and Differential Evolution are intentionally
not run automatically in this notebook because they can require many thousands
of additional simulator evaluations.

After confirming the BiLSTM SIR/SEIR pipeline, they can be run through:

```r
baseline_calibrators()
compare_calibrators()
```

using the same held-out test curves.

The detailed SEIR/HPC versions of those methods remain in:

```text
examples/seir_reference_comparison/
```
